National Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Beliefs

### 2020-11-25 Manuscript_Bayesian.Rmd
added expectations. need to fix all figures and rewrite results.

```{r runanalysis, include = FALSE}
source('short.R')
```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three credible challenges.

@@

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.50, se =.03) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.48, se =.03) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.64,se =.03) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.66, se =.05). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory is lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 4.86, se =.02) and lower among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ = 4.91, se =.02) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 4.94,se =.02) and the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.02, se =.02) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y}$ = 6.67, se =.03, $\Delta \hat{y} = 0.17$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} = 6.69, se = 0.02, \Delta \hat{y} = .21$). To place this effect in practical perspective, such growth in personal well-being is about twice as large as the expected benefit of being employed and about a third as large as the expected benefit from having a romantic partner. By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force remained stable across the eight year interval $\hat{y} = 6.66, se = 0.03, \Delta \hat{y} = .02$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  6.70, se = 0.04, \Delta \hat{y} = .04$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 5.01, se = 0.02, \Delta \hat{y} = 0.07$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.03, se = 0.02, \Delta \hat{y} = 0.12$). In practical terms, such growth is equivalent to 1.5 times the expected gain to life satisfaction from being employed and almost a third benefit to life satisfaction from having a romantic partner. By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force remained stable over the eight year interval $\hat{y} =  4.94, se =0.02, \Delta \hat{y} = 0$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force was constant $\hat{y} =  5.00, se = 0.03, \Delta \hat{y} = -.02$) (See Figure 3).

>>>>

# ```{r runanalysis, include = FALSE}
# source('short.R')
# ```
The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Most previous attempts to quantify the relationship between these two domains find a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three credible challenges.

@@

```{r fig.cap = "Combination Coefficient Plots for personal well-being and life satisfaction longitudinal growth models"}
knitr::include_graphics("/Users/joseph/Documents/GitHub/nzavs_spiritbeliefWellbeing/Figs/USE_2_coefficientplots-1.png")

@@

1. At the outset of the study, in 2010, the expected subjective well-being as measured using the personal well-being Inventory was expected to be lower among those who believed in Spirit or Life Force but not a God ($\hat{y}$ =  6.78 [6.74, 6.82]) and lower among those who believed in both a Spirit/Life-Force and a God ($\hat{y}$ =  6.75 [6.72, 6.78]) in comparison both to the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 6.91 [6.86, 6.95]) and to the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 6.88 [6.81, 6.96]). (See Figure 2)  Similarly,the expected subjective well-being as measured using the life satisfaction Inventory was lowest in 2010 among those who believed in Spirit or Life Force but not a God ($\hat{y}$ = 5.07 [5.04, 5.10]). Life satisfaction was similar among those who believed in both a Spirit/Life-Force and a God  ($\hat{y}$ =  5.11 [5.08, 5.13]) in comparison both with the skeptical population who neither believed in a God nor a life force ($\hat{y}$ = 5.12 [5.08, 5.15]). Life satisfaction was highest among the population that believed in a God but was skeptical about belief in a Spirit or Life Force ($\hat{y}$ 5.19 [5.14, 5.25]) (see Figure 3)

2. However, over time, belief in a Spirit or Life Force predicts a trajectory of growth in personal well-being. In 2018, personal well-being among who believe in Spirit or Life Force but not a God had grown to ($\hat{y} = 7.13$ [7.09, 7.16], $\Delta \hat{y} = 0.35$). Among those who believed in both a Spirit/Life-Force and a God, personal well-being had grown to ($\hat{y} =  7.14$, [7.11, 7.16], $\Delta \hat{y} = 0.39$). By contrast, personal well-being among the skeptical population who neither believed in a God nor a Life Force grew less steeply across the eight year interval $\hat{y} = 7.15 [7.12, 7.19]$, $\Delta \hat{y} = 0.24$). By the same token, personal well-being among those who believe in a God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} = 7.15$, [7.09, 7.21] $\Delta \hat{y} = 0.27$). Similarly, over time, belief in a Spirit or Life Force predicts a trajectory of growth in life satisfaction. In 2018, the expected effect of spiritual beliefs on life satisfaction among those who believe in Spirit or Life Force but not a God had grown to ($\hat{y} =  5.30$ [5.28, 5.33], $\Delta \hat{y} = 0.23$). In practical terms, such growth is equivalent to the expected gain in life satisfaction from being employed and a sixth as large as the expected benefit to life satisfaction from having a romantic partner. In practical termss, the rate of increase in personal wellbeing among those with Spirit/Life force beliefs was about one and half times greater than it was among those who lacked such beliefs (See Figure 2). Among those who believed in both a Spirit/Life-Force and a God, life satisfaction had grown to ($\hat{y} = 5.33 [5.31, 5.35]$, $\Delta \hat{y} = 0.22$). By contrast, life satisfaction among the skeptical population who neither believed in a God nor a Life Force lagged behind over the eight year interval $\hat{y} =  5.24 [5.21, 5.27], \Delta \hat{y} = 0.12$). Relatedly, life satisfaction in the population that believe in God but are skeptical about a Spirit/Life Force grew less steeply $\hat{y} =  5.32 [5.27, 5.36], \Delta \hat{y} = 0.13$). In practical terms, the rate of increase in life satisfaction among those with Spirit/Life force beliefs was twice as large (See Figure 3).

### 2020-11-25 Supplement_And_Analysis.Rmd
added expectations. need to fix all figures and rewrite results.

---
title             : "Supplement:Analysis for Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Supplement 1"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

authornote: |
  1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

@@

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

@@

```{r, echo = TRUE}
# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
d<-data_read_mymachine()

@@

```{r latex_data_table, include=FALSE}
# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
demographic_table_latex(d_3)
```

@@

# function to count waves for participants
count_waves_participants(d_3)
count_waves_participants(d_4)
count_waves_participants(d_5)
count_waves_participants(d_6)
count_waves_participants(d_7)
count_waves_participants(d_8)
count_waves_participants(d_9)
```

@@

```{r robust to years and beliefs only, cache = TRUE, dependson=dat_3,cache =TRUE}
library(lme4)
pwi_formula_simple <- 'PWI ~  years * Beliefs + (1+ Beliefs|Id)'
ls_formula_simple <- 'LIFESAT ~  years * Beliefs + (1+ Beliefs|Id)'
mod.1aa <-lmer(pwi_formula_simple, data= d_3)
mod.2aa <-lmer(ls_formula_simple, data= d_3)

@@

```{r prediction plot of ML longitudinal model, cache =TRUE, dependson=mod.1aa}
```{r bayesian_pwi_predict, cache =TRUE}
graph_predictions(mod.1aa, "Predicted Values of Personal Wellbeing")
```

@@

```{r table of saturated models, cached =TRUE}
sjPlot::tab_model(mod.1a,mod.2a)
```

@@

```{r plot_coefficientplots_LMER_FULL, cache =TRUE}
g1/g2
```

Next, recall we want to check whether inference is  robustness to the researcher degree of freedom in selecting the number of waves.
```{r run all the models with different ns, cache=TRUE}
mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
mod.2_3 <-lmer (ls_formula_simple, data= d_4)

@@

```{r setmodelformula}
forumula_bayesPWI <- bf(PWI ~ years * Beliefs + (1 + Beliefs|Id))
forumula_bayesLIFESAT <- bf(LIFESAT ~  years * Beliefs + (1 + Beliefs|Id))
```
```{r pwi_bayesmodel}
# PWI
mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds") # note these are very large files
summary(mod.1_b)
#saveRDS(mod.1_b,"backup_mod.1_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
```
```{r lsbayesmodel}
# LIFE SAT
mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") # note these are very large files
# saveRDS(mod.2_b,"backup_mod.2_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
```

Tables
```{r lifesat bayesian model table:pwi, cache =TRUE}
tab_model(mod.1_b)
```

@@

```{r mixing of sd_pwi}
plot(mod.1_b, pars ="b_") #requires stronger priors
```

@@

```{r posterior predictive checks bayes:pwi, cache=TRUE}

pp_1<-NULL
pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")

pp_2<-NULL
pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")

# plot
pp_1 + pp_2
```

@@

Prediction plots
```{r bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
graph_predictions(mod.1_b, "Predicted Values of Personal Wellbeing")
```
```{r bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
```

>>>>

---
title             : "Supplement:Analysis for Longitudinal Evidence for Growth in Subjective Wellbeing from Spiritual Belief"
shorttitle        : "Supplement 1"
author: 
  - name          : "Benjamin Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "520 Power Plant Circle Apartment 105 Winston-Salem, NC 27101 USA"
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph A. Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"
authornote: |
  1 Wake Forest School of Medicine 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

@@

```{r setup, include = FALSE}
#library("papaja")
#r_refs("r-references.bib")
```

@@

```{r, read_data, cache=TRUE}
# custom function for my machine to read data. you'll need to read data to your machine. tweak function in the `funs.R' script for your machine
d<-data_read_mymachine()

@@

```{r latex_data_table, include=FALSE}
# you might need to install the dev verion.//remotes::install_github("tysonstanley/furniture")
#demographic_table_latex(d_3)
```

@@

# function to count waves for participants
count_waves_participants(d_3)
count_waves_participants(d_4)
count_waves_participants(d_5)
count_waves_participants(d_6)
count_waves_participants(d_7)
count_waves_participants(d_8)
```

@@

```{r robust to years and beliefs only, cache = TRUE, dependson=d_3,cache =TRUE}
library(lme4)
pwi_formula_simple <- 'PWI ~  years * Beliefs + (1+ Beliefs|Id)'
ls_formula_simple <- 'LIFESAT ~  years * Beliefs + (1+ Beliefs|Id)'
mod.1aa <-lmer(pwi_formula_simple, data= d_3)
mod.2aa <-lmer(ls_formula_simple, data= d_3)

@@

```{r prediction plot of ML longitudinal model, cache =TRUE}
```{r bayesian_pwi_predict, cache =TRUE}
graph_predictions(mod.1aa, "Predicted Values of Personal Wellbeing")
```

@@

```{r table of saturated models, cached = TRUE}
sjPlot::tab_model(mod.1a,mod.2a)
```

@@

```{r plot_coefficientplots_LMER_FULL, cache = TRUE}
g1/g2
```

Next, recall we want to check whether inference is  robustness to the researcher degree of freedom in selecting the number of waves.
```{r run all the models with different ns, cache = TRUE}
mod.1_3 <-lmer (pwi_formula_simple, data= d_4)
mod.2_3 <-lmer (ls_formula_simple, data= d_4)

@@

```{r setmodelformula}
forumula_bayesPWI <- bf(PWI ~ years * Beliefs + (1 + Beliefs|Id))
forumula_bayesLIFESAT <- bf(LIFESAT ~  years * Beliefs + (1 + Beliefs|Id))
```
```{r pwi_bayesmodel}
# PWI
mod.1_b <-brm(forumula_bayesPWI, data= d_3, file ="mod.1_b.rds") # note these are very large files
summary(mod.1_b)
#saveRDS(mod.1_b,"backup_mod.1_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
rm(mod.1_b)
```

```{r lsbayesmodel}
# LIFE SAT
mod.2_b <-brm (forumula_bayesLIFESAT, data= d_3,cores = 8, file ="mod.2_b.rds") # note these are very large files
# saveRDS(mod.2_b,"backup_mod.2_b") this is excessive, but we save another backup to avoid running this model again in case of SNAFU.
```

Tables:
```{r lifesat bayesian model table:pwi, cache =TRUE}
tab_model(mod.1_b)
```

@@

```{r mixing of sd_pwi}
plot(mod.1_b, pars ="b_") 
```

@@

# ```{r posterior predictive checks bayes:pwi, cache=TRUE}
# # cannot run on my laptop
# # pp_1<-rstanarm::pp_check(mod.1_b) + ylab("Personal Wellbeing")
# # pp_2<-rstanarm::pp_check(mod.2_b) + ylab("Life Satisfaction")
# # 
# # # plot
# pp_1 + pp_2
#```

@@

Predictions and plots of them. First we get the predictions.

```{r}
preds_1b<-get_predictions(mod.1_b)
preds_1b
```

```{r}
preds_2b<-get_predictions(mod.2_b)
preds_2b
```

```{r bayesian_pwi_predicted, cache=TRUE, dependson= mod.1_b}
graph_predictions(mod.1_b, "Predicted Values of Personal Wellbeing")
```
```{r bayesian_lifesat_predicted, cache=TRUE, dependson= mod.2_b}
graph_predictions(mod.2_b, "Predicted Values of Life Satisfaction")
```
